A New Method of Modelling Tuneable Lasers with Functional Composition

### 2019-08-29 Thesis/Model/Model.tex
Updated for binding

\chapter{A New Model}
\label{chap:linear}
In this chapter we shall derive our new model and solve it analytically in the linear case. To accomplish this we shall use the ideas of the previous functional models \cite{cutler, siegman, kuizenga1970a, kuizenga1970b, kuizenga1970, martinez1984, martinez1985, burgoyne2014}. To alleviate some of the pitfalls mentioned in the previous chapter, we shall include all five processes involved in the modification of the pulse within the laser cavity (gain, nonlinearity, loss, dispersion, and modulation). In addition to this, the functional operations associated with each component will be derived from \eqref{eq:nlse}, with the exception of the modulation in which we consider the exact functional form to be determined by the laser operator. \\

\section{Components}
\label{sec:comp}
We shall begin our analysis with the derivation of the functional operators for the five components. \\

@@

Despite assuming a Gaussian modulation, the solution presented in Section \ref{sec:linear} can be generalized to any modulation function in $L^2(\mathbb{R})$ since Gaussians in fact span $L^2(\mathbb{R})$ as shown in Appendix \ref{chap:gauss}. \\

@@

\section{Solution to the Linear Model}
\label{sec:linear}
In reality the nonlinearity parameter, $b$, is of order unity by \eqref{eq:ndparam}, but we shall start our investigation with the simpler case of $b = 0$---which shall be referred to as the linear model. In this case, a solution can be found analytically. It is expected the solution will take the form of a Gaussian. There are a few reasons for this; the solution to the previous models were Gaussian \cite{cutler, siegman, kuizenga1970a, martinez1984, martinez1985} (see Section \ref{chap:meml}), the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn} dispersion will not alter the envelope. \\

\begin{figure}[tbp]
\centering
\input{./Figures/Sample_Gauss}
\caption{Sample realization of \eqref{eq:A0}.}
\label{fig:samplegauss}
\end{figure}

@@

In equilibrium it must be that $A_0 = A_4 \textrm{e}^{-i \phi} = \mathcal{L}(A_0) \textrm{e}^{- i \phi}$ so that after a single loop of the cavity the variance, chirp, and envelope remain unchanged. More explicitly, this gives three conditions:
\begin{subequations}
\begin{align}
\label{eq:energycond}
1 &= \left( \frac{W(a E \textrm{e}^E)}{E} \right)^{1/2} h \zeta, \\
\label{eq:varcond}
\frac{1}{\sigma^2} &= \frac{1}{\widetilde{\sigma}^2} + 1, \\
\label{eq:chirpcond}
\frac{C}{\sigma^2} &= \frac{\widetilde{C}}{\widetilde{\sigma}^2}.
\end{align}
\label{eq:linsys}
\end{subequations}

\subsection{Spread of the Pulse Due to Dispersion}
Each of these processes has a relatively straight forward effect, with the exception of dispersion.
\input{./Appendices/Dispersion}

@@

\subsection{Asymptotic Expansion of the Variance}
\begin{figure}[tbp]
\centering
\input{./Figures/Asympt}
\caption{Plot of \eqref{eq:asymp} and \eqref{eq:asymp2} to highlight the structure.}
\label{fig:asympt}
\end{figure}
While \eqref{eq:equilvar} is useful, so too is the asymptotic behaviour. The asymptotic expansion provides additional information about the structure that the full analytic expression cannot. Moveover, the expansion provides more detail about the limits, as well has the rate of convergence. Lastly, the expansion allows for simplified expressions within their domains that are much less unwieldy than those found in the previous subsection. \\

@@

With the linear model fully solved, we shall now turn our attention to the rich structure added by the full nonlinear model. \\

>>>>

\chapter{A New Model}
\label{chap:linear}
In this chapter we shall derive our new model and solve it analytically in the linear case. To accomplish this we shall use the ideas presented in the previous chapter of the prior functional models \cite{cutler, siegman, kuizenga1970a, kuizenga1970b, kuizenga1970, martinez1984, martinez1985, burgoyne2014}. To alleviate some of the pitfalls mentioned in the previous chapter, we shall include all five processes involved in the modification of the pulse within the laser cavity (gain, nonlinearity, loss, dispersion, and modulation). In addition to this, the functional operations associated with each component will be derived from \eqref{eq:nlse}, with the exception of the modulation in which we consider the exact functional form to be determined by the laser operator.

\section{Components}
\label{sec:comp}
We shall begin our analysis with the derivation of the functional operators for the five components.

@@

Despite assuming a Gaussian modulation, the solution presented in Section \ref{sec:linear} can be generalized to any modulation function in $L^2(\mathbb{R})$ since Gaussians in fact span $L^2(\mathbb{R})$ as shown in Appendix \ref{chap:gauss}.

@@

\section{Solution to the Linear Model}
\label{sec:linear}
\begin{figure}[tbp]
\centering
\input{./Figures/Sample_Gauss}
\caption{Sample realization of \eqref{eq:A0}.}
\label{fig:samplegauss}
\end{figure}
In reality the nonlinearity parameter, $b$, is of order unity by \eqref{eq:ndparam}, but we shall start our investigation with the simpler case of $b = 0$---which shall be referred to as the linear model. In this case, a solution can be found analytically. It is expected the solution will take the form of a Gaussian. There are a few reasons for this; the solution to the previous models were Gaussian \cite{cutler, siegman, kuizenga1970a, martinez1984, martinez1985} (see Section \ref{chap:meml}), the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn} dispersion will not alter the envelope. \\

@@

In equilibrium it must be that $A_0 = A_4 \textrm{e}^{-i \phi} = \mathcal{L}(A_0) \textrm{e}^{- i \phi}$ so that after a single loop of the cavity the variance, chirp, and envelope remain unchanged. More explicitly, this gives three conditions:
\begin{subequations}
\begin{align}
\label{eq:energycond}
1 &= \left( \frac{W(a E \textrm{e}^E)}{E} \right)^{1/2} h \zeta, \\
\label{eq:varcond}
\frac{1}{\sigma^2} &= \frac{1}{\widetilde{\sigma}^2} + 1, \\
\label{eq:chirpcond}
\frac{C}{\sigma^2} &= \frac{\widetilde{C}}{\widetilde{\sigma}^2}.
\end{align}
\label{eq:linsys}
\end{subequations}
\vspace{-5mm}
\subsection{Spread of the Pulse Due to Dispersion}
Each of these processes has a relatively straight forward effect, with the exception of dispersion.
\input{./Appendices/Dispersion}

@@

\subsection{Asymptotic Expansion of the Variance}
While \eqref{eq:equilvar} is useful, so too is the asymptotic behaviour. The asymptotic expansion provides additional information about the structure that the full analytic expression cannot. Moveover, the expansion provides more detail about the limits, as well has the rate of convergence. Lastly, the expansion allows for simplified expressions within their domains that are much less unwieldy than those found in the previous subsection. \\

@@

With the linear model fully solved, we shall now turn our attention to the rich structure added by the full nonlinear model.

### 2019-08-29 Thesis/Nonlinear/Nonlinear.tex
Updated for binding

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique weakly bi-modal shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, and by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, we recover this linear response for moderate values of $T$. However, for $|T| > 1$ the chirp begins to saturate. This is consistent with the experimental results \cite{chen, rothenberg, tomlinson}. \\

@@

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. In this nonlinear case we allow the initial pulse to pass through each component 100 times instead of 40 in the linearized case. This choice can be justified by Figure \ref{fig:roc}. The pulse reaches a state in which the error no longer decreases within typically 10--20 iterations. In the linearized case we chose to iterate 40 times since this is about double the number of iterations expected. Furthermore, in the nonlinear case we extend this number further to 100 in case the nonlinearity increases the number of iterations required to converge. The energy of the pulse after 100 iterations is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$ and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy with very large derivatives. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary between periodic and aperiodic behaviour appears to become periodic (in the parameter space) as $b$ increases around $s = 0.3$. These features are further highlighted in Figure \ref{fig:cartoon} by disregarding the energy and only examining whether the pulse is invariant after additional iterations. \\

@@

\subsubsection{\glslink{spm}{Self-Phase Modulation} and Wave Breaking}
\glsreset{spm}

The noise in the energy exhibited for moderate to large values of $b$, and small values of $s$ is a phenomenon called \emph{wave breaking} \cite{agrawal2013, anderson, finot, rothenberg, tomlinson}. Wave breaking is not limited to just optics; wave breaking occurs in areas such as plasmas, transmission lines, and fluid dynamics \cite{rothenberg}. Wave breaking occurs because the pulse begins to interfere with itself in a way called \gls{spm} \cite{agrawal2002, agrawal2013, becker}. \gls{spm} occurs because the index of refraction is intensity dependent \cite{agrawal2002, becker, rothenberg, silfvast}, which leads to additional chirp across the pulse \cite{agrawal2013, anderson, rothenberg, silfvast}. This in turn causes higher order frequencies to be injected into the pulse \cite{agrawal2013, anderson}, as we saw in Figure \ref{fig:nlstable}. These high frequencies compound with each trip around the cavity becoming parasitic very quickly---Figure \ref{fig:break} highlights this. Notice that the difference between Figure \ref{fig:nlstable} and Figure \ref{fig:break} is a difference in $b$ of $0.05$---just enough to cross the boundary---this difference could be attributed to adding a few centimetres more of fibre between the gain and output coupler. The left figures show the pulse after 11 trips around the cavity; in the Fourier transform it is clear that the contributions from higher frequencies has increased. We obtain comparable results as in the experiments \cite{anderson, rothenberg}. Additionally, the chirp starts losing its linearity causing it to start becoming unstable; the nature of this instability is again in agreement with the experiments \cite{anderson, rothenberg}. The parasitic nature of the high frequency contributions is evident by examining the right figures. After five additional trips around the cavity, the envelope of the pulse is much more rippled, and the real and imaginary parts become incoherent. Moreover, the Fourier transform has no clear structure and has essentially become noise. The chirp has grown to be highly oscillatory and unstable. Once the pulse has reached a state such as this, the envelope, Fourier transform, and chirp never reach a steady equilibrium state. \\

@@

We now have the data to make a more meaningful estimate of the error between iterations. To compute this composite error we take the minimum of the error from each of these calculations:
\begin{align}
\textrm{E}_c = \min_\Delta \left\{ \textrm{E}_\Delta \right\}.
\label{eq:errorcomp}
\end{align}
The intriguing structure of the composite error is shown in Figure \ref{fig:errorcomp}. Compared with the error from Figure \ref{fig:error} the lower right region is much more well behaved, that is, the error is a few orders of magnitude lower. A peculiar trait of the band between the unstable and stable regions is that within this area the pulse is in a quasi-stable state---the pulse is stable and has reached a sort of equilibrium, however, the envelope of the pulse has small variations with no clear period. Additionally, there appears to be no periodicity of the envelope within the wave breaking region, further supporting the claim that the pulse here is completely unstable. \\

>>>>

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique weakly bi-modal shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, and by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, we recover this linear response for moderate values of $T$. However, for $|T| > 1$ the chirp begins to saturate. This is consistent with the experimental results \cite{chen, rothenberg, tomlinson}.

@@

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. In this nonlinear case we allow the initial pulse to pass through each component 100 times instead of 40 in the linearized case. This choice can be justified by Figure \ref{fig:roc}. The pulse reaches a state in which the error no longer decreases within typically 10--20 iterations. In the linearized case we chose to iterate 40 times since this is about double the number of iterations expected. Furthermore, in the nonlinear case we extend this number further to 100 in case the nonlinearity increases the number of iterations required to converge. The energy of the pulse after 100 iterations is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$ and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy with very large derivatives. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary between periodic and aperiodic behaviour appears to become periodic (in the parameter space) as $b$ increases around $s = 0.3$. These features are further highlighted in Figure \ref{fig:cartoon} by disregarding the energy and only examining whether the pulse is invariant after additional iterations.

@@

\subsubsection{\glslink{spm}{Self-Phase Modulation} and Wave Breaking}
\glsreset{spm}
The noise in the energy exhibited for moderate to large values of $b$, and small values of $s$ is a phenomenon called \emph{wave breaking} \cite{agrawal2013, anderson, finot, rothenberg, tomlinson}. Wave breaking is not limited to just optics; wave breaking occurs in areas such as plasmas, transmission lines, and fluid dynamics \cite{rothenberg}. Wave breaking occurs because the pulse begins to interfere with itself in a way called \gls{spm} \cite{agrawal2002, agrawal2013, becker}. \gls{spm} occurs because the index of refraction is intensity dependent \cite{agrawal2002, becker, rothenberg, silfvast}, which leads to additional chirp across the pulse \cite{agrawal2013, anderson, rothenberg, silfvast}. This in turn causes higher order frequencies to be injected into the pulse \cite{agrawal2013, anderson}, as we saw in Figure \ref{fig:nlstable}. These high frequencies compound with each trip around the cavity becoming parasitic very quickly---Figure \ref{fig:break} highlights this. Notice that the difference between Figure \ref{fig:nlstable} and Figure \ref{fig:break} is a difference in $b$ of $0.05$---just enough to cross the boundary---this difference could be attributed to adding a few centimetres more of fibre between the gain and output coupler. \\

The left figures show the pulse after 11 trips around the cavity; in the Fourier transform it is clear that the contributions from higher frequencies has increased. We obtain comparable results as in the experiments \cite{anderson, rothenberg}. Additionally, the chirp starts losing its linearity causing it to start becoming unstable; the nature of this instability is again in agreement with the experiments \cite{anderson, rothenberg}. The parasitic nature of the high frequency contributions is evident by examining the right figures. After five additional trips around the cavity, the envelope of the pulse is much more rippled, and the real and imaginary parts become incoherent. Moreover, the Fourier transform has no clear structure and has essentially become noise. The chirp has grown to be highly oscillatory and unstable. Once the pulse has reached a state such as this, the envelope, Fourier transform, and chirp never reach a steady equilibrium state.

@@

We now have the data to make a more meaningful estimate of the error between iterations. To compute this composite error we take the minimum of the error from each of these calculations:
\begin{align}
\textrm{E}_c = \min_\Delta \left\{ \textrm{E}_\Delta \right\}.
\label{eq:errorcomp}
\end{align}
The intriguing structure of the composite error is shown in Figure \ref{fig:errorcomp}. Compared with the error from Figure \ref{fig:error} the lower right region is much more well behaved, that is, the error is a few orders of magnitude lower. A peculiar trait of the band between the unstable and stable regions is that within this area the pulse is in a quasi-stable state---the pulse is stable and has reached a sort of equilibrium, however, the envelope of the pulse has small variations with no clear period. Additionally, there appears to be no periodicity of the envelope within the wave breaking region, further supporting the claim that the pulse here is completely unstable.
